Revealing Transient Strain in Geodetic Data with the Radial Basis Function Finite Difference Method

### 2017-06-06 HinesHetlandGPR.tex
added more

\section{Method}\label{sec:Method}
In this section, we describe how GNSS displacement observations are used to identify transient crustal strain rates, which we denote as $\dot{\bm{\varepsilon}}(z)$, where $z$ is the ordered pair $(x,t)$, $x$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We consider $\dot{\bm{\varepsilon}}$ to be spatially and temporally coherent deviations from the steady rate of strain accumulation from plate tectonics. We determine $\dot{\bm{\varepsilon}}$ by first identifying transient displacements, $\bm{u}(z)$, which we then spatially and temporally differentiate.  As we will show in Section X, estimates of $\dot{\bm{\varepsilon}}$ turn out to be more effective at illuminating geophysical signal than estimates of $\bm{u}$ or $\dot{\bm{u}}$.  We make a prior assumption that each component of $\bm{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(z) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(z,z')$ is a covariance function indicating how we expect $u_i(z)$ to covary with $u_i(z')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We then drop the component subscripts with the understanding that the same analysis is being repeated to estimate the east, north, and vertical components of $\bm{u}$. We further assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((x,t),(x',t')\right) = X(x,x')T(t,t').
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in the next section.

@@

Our method for detecting outliers is based on the data editing algorithm described in \citet{Gibbs2011}.  We again treat $\bm{d}_*$ as a sample of $\bm{d}$, and we assume some prior covariance function for $u$. The choice for $u$ here does not need to be the same as that used in Section X, and a conservative approach would be to choose a $u$ that is more flexible (e.g., shorter characteristic length-scales and higher variance).  We let $\epsilon$ be uncorrelated noise which has standard deviation given by the formal data uncertainties, $\bm{\sigma}$. We consider outliers to be observations which are abnormally inconsistent with the random vector $u(\bm{y}) + \bm{P}\bm{a}$.  With this criterion, we takes into account the spatial and temporal relationship between the data, allowing us to identify anomalous observations which may not be immediately apparent based on inspection of individual timeseries. To begin the algorithm, we let $\Omega$ be the index set of non-outliers in $\bm{d}_*$ and initiate it with all $n$ indices. This algorithm is iterative, and for each iteration we calculate the residual vector
\begin{align}\label{eq:Residual}
\bm{r} &= \frac{\bm{d}_* - \mathrm{E}\left[(u(\bm{y}) + \bm{P}\bm{a})|\tilde{\bm{d}}_* \right]}{\bm{\sigma}} \\
       &= \frac{1}{\bm{\sigma}}\left(\bm{d}_*  - 
\left[ 
\begin{array}{cc}
C_u(\bm{y},\tilde{\bm{y}}) & \bm{P} \\
\end{array}
\right]
\left[
\begin{array}{cc}
C_u(\tilde{\bm{y}},\tilde{\bm{y}}) + \mathrm{diag}(\tilde{\bm{\sigma}}^2) & \tilde{\bm{P}} \\
\tilde{\bm{P}}^T  & \bm{0} \\
\end{array}
\right]^{-1}
\left[
\begin{array}{c}
\tilde{\bm{d}}_* \\
\bm{0} \\
\end{array}
\right] 
\right),
\end{align}
where the tilde indicates that only elements corresponding to indices in $\Omega$ are retained (e.g., $\tilde{\bm{y}} = (y_i)_{i\in\Omega}$). We then update $\Omega$ to be
\begin{equation}\label{eq:Update}
\Omega = \{i : |r_i| < \eta \tilde{\bm{r}}_{rms}\}
\end{equation} 
where $\tilde{\bm{r}}_{rms}$ is the root-mean-square of $\tilde{\bm{r}}$ and $\lambda$ is an outlier tolerance. We use $\lambda=4$ in this study, which seems to accurately identify outliers without unnecessarily decimating the data. Iterations continue until the new $\Omega$ is equal to the previous $\Omega$ (i.e., no new outliers are detected).

@@

GIVE A DEMONSTRATION HERE

\section{Application Cascadia Slow Slip Events}\label{sec:Cascadia}
We demonstrate the method for estimating transient strain described in the previous section by applying it to GNSS data in Cascadia. We are mainly interested in identifying transient strain resulting from slow slip events (SSEs) \citep[e.g.,][]{Dragert2001}. We show that Gaussian processes are an effective tool for (1) detecting SSEs and (2) understanding how the strain from SSEs are contributing to the strain accumulated over the seismic cycle. Our primary motivation for detecting SSEs is because there have been several documented cases of SSEs triggering megathrust earthquakes \citep{Roeloffs2006}.  In Cascadia, there is less of a need to detect SSEs with GNSS data because the SSEs are associated with a spike in seismic tremor, which is actively monitored by the Pacific Northwest Seismic Network (PNSN) \citep{Wech2010}. However, SSEs have been observed in other subduction zones without any associated tremor \citep{Schwartz2007}, and so there is still utility in developing method for detecting SSEs with GNSS data. Furthermore, by comparing our estimated transient strain in Cascadia with the tremor catalog, we can verify the accuracy of method for detecting SSEs.

@@

To determine the appropriate hyperparameters for the FOGM model, $\alpha$ and $\beta$, we first select 38 stations in Washington and Oregon which are east of $121^\circ$ W, where we can reasonably expect no transient signal from SSEs (i.e. $u = 0$). We then find $\alpha$ and $\beta$ for each station timeseries with the Restricted Maximum Likelihood (REML) method \cite[e.g.,][]{Harville1974,Cressie1992} and Hines 2017. The REML method finds the hyperparameters which maximize the probability of drawing $\bm{d}_*$ from $\bm{d}$. We use the REML method over the maximum likelihood (ML) method \citep[e.g.,][]{Langbein1997} because the REML method accounts for the improper prior that we assigned to $\bm{a}$ without biasing the hyperparameter estimates. The distribution of inferred $\alpha$ and $\beta$ are shown in Figure X. The amplitude of FOGM noise for the east and north components is notable low and are clustered around 0.5 mm/yr$^{0.5}$. The corresponding estimates of $\alpha$ tend to cluster around 0 yr$^{-1}$, suggesting that Brownian motion would have also been an appropriate noise model.  The amplitude of FOGM noise for the vertical component is significantly larger with a median value of 13.8 mm/yr$^{0.5}$. The inferred values for alpha are higher for the vertical component with a median value of 1.87 yr$^{-1}$. In Figure X, we use the median values of $\alpha$ and $\beta$ to generate two random samples of FOGM noise. The samples span seven years and over these seven years the east and north samples drift by about 1 mm. In the context of detecting SSEs, which produce several mm's of surface displacement on the timescale of weeks, the estimated FOGM noise for the east and north component is negligible. In contrast, the estimated FOGM noise for the vertical component is significantly larger than the signal we would expect from SSEs. We suspect that the higher amplitude for the FOGM model in the vertical component is accommodating for deficiencies in our rather simple seasonal model. Based on this analysis, we henceforth ignore temporally correlated noise in the east and north component because of its low amplitude, and we do not estimate $u$ for vertical deformation because of the low signal to noise ratio.

@@

\subsection{Transient displacement model}
We next discuss our prior model for transient displacements, and, specifically, our choice for the covariance functions $X(x,x')$ and $T(t,t')$. For $X$, we use the squared exponential (SE) covariance function,
\begin{equation}\label{eq:SE}
X(x,x') = \alpha^2 \exp\left(\frac{-||x - x'||_2^2}{2 \beta^2}\right).
\end{equation}
The SE covariance function is commonly used in Kriging \citep[e.g,][]{Cressie1992} and Gaussian process regression \citep[e.g.,][]{Rasmussen2006}.  The SE is a valid (i.e. positive definite) covariance function for any number of spatial dimensions, and it describes an isotropic Gaussian processes with realizations that are infinitely differentiable. In terms of geodetic applications, \citet{Kato1998} and \cite{El-Fiky1999} demonstrated that the SE is an appropriate covariance model for describing long-term tectonic strain rates in Japan.

We consider three potential covariance functions to describe the temporal covariance of $u$.  First, we consider the one-dimensional SE covariance function, simply because the SE describes a good general purpose Gaussian process. Second, we consider integrated Brownian motion (IBM). IBM has zero mean and its covariance function can be found by integrating the covariance function for Brownian motion as
\begin{align}\label{eq:IBM}
T(t,t') &= \int_0^t \int_0^{t'} \alpha^2 \min(\tau,\tau') \,d\tau'\,d\tau \\
        &= \frac{\alpha^2}{2}\min(t,t')^2 \left(\max(t,t') - \frac{1}{3}\min(t,t')\right), \ \ \ t,t' \geq 0.
\end{align}
IBM has been used, in the context of Kalman filtering, as a nonparametric model for the time dependence of geophysical signals \citep[e.g.,][]{Segall1997,McGuire2003,Ohtani2010,Hines2016}. It should be noted that IBM has a reference time at which the Gaussian process is exactly zero. For some geophysical signals, such as postseismic deformation, it is appropriate to have this reference time. However, if we are interesting in detecting transient events, where there is no known start time, then the IBM model may not be appropriate and an isotropic Gaussian process should be preferred. Alternatively we can assume that the start date for the IBM process is long before the first data obvservation; however this can introduce numerical difficulties because the covariance matrix will become ill-conditioned. Third, we consider the Wendland class of covariance functions \citep{Wendland2005}. Wendland functions are positive definite in $\mathbb{R}^d$, and they describes an isotropic Gaussian process with realizations that can be differentiated $k$ times. The form of the covariance function depends on the choice of $d$ and $k$. Wendland covariance functions have compact support, which is a particularly useful property because we can exploit the sparsity of the the corresponding covariance matrices. To describe the temporal covariance of $u$, we only require that $d=1$ and $k=1$. The corresponding Wendland covariance function is 
\begin{equation}\label{eq:Wendland}
T(t,t') = \alpha^2\left(1 - \frac{|t - t'|}{\beta}\right)^3_+ \left(\frac{3|t - t'|}{\beta} + 1\right), 
\end{equation}
where
\begin{equation}
(t)_+ = 
\begin{cases}
t, \ \ \ t > 0 \\
0, \ \ \ \mathrm{otherwise}.
\end{cases}
\end{equation}

We next find the optimal hyperparameters for $X$ and each of the three candidate functions for $T$. We want to find the hyperparameters which best describes the spatial and temporal statistics of ground deformation resulting from SSEs. We focus on describing the SSEs in the Puget Sound region, which regularly occur with a frequency of about 14 months \citep{Brudzinski2007}.

To describe these events we use the set of GNSS station shown in Figure X. We use the GNSS data that has been made publicly available through UNAVCO \citep{Herring2016}. We first remove timeseries jumps based on the compiled GNSS offsets from blewitt. We then use the data editing algorithm to remove outliers. We find the data editing algorithm is particularly useful at removing outliers for stations at higher altitudes, which can be adversely affected by snow \citep{Lisowski2008}. For our data editing algorithm we use X for the prior and we give an example in figure X.

We use the REML method to find the hyperparameters that best describe SSE events which occurred since 2010, corresponding to the time that coverage improved due to the PBO network.  We identify the times of seven SSEs using tremor records from \cite{Wech2010}, and we break the spatio-temporal GNSS data into four month long subsets roughly centered on the time of each SSE. For each data subset use the REML method to find the hyperparameters for $T$ and $X$ that best describe deformation from the SSE. Our motivation analyzing subsets of the data is two-fold: (1) it keeps the analysis computationally tractible and (2) the distribution of estimated hyperparameters will indicates whether we can use a single covariance model to describe all SSEs. Our hyperparameter estimates are summarized in Table X.

We next want to identify which covariance function for $T$ best describes the SSEs. We could use the REML likelihoods to compare different models, similar to \citep{Langbein2004}. However, the maximum likelihoods do not vary by more than a factor of X between the models, which does not definitively identify the best model. We then compare the posterior estimates $\hat{u}$ for each model to the observations to see if any model is noticeably better at describing the data. We show $\hat{u}$ using the three different function for $T$ in Figure X. It is clear that neither model offer any distinct improvement over the others. For most stations $\hat{u}$ appears to accurately capture the deformation from SSEs; however, for some stations which are closer to Puget Sound such as X and X, the rate of deformation is underestimated by the best fitting model. We could have manually specified a prior. We choose the Wendland function to describe $T$ because evaluating $T$ will yield sparse covariance matrices. We utilize the sparsity of the covariance matrices with the sparse matrix library CHOLMOD \citep{Chen2008}, which significantly expands the amount of data which we are able to tractably analyze with the method described in Section X.

\subsection{Transient Strain Rates}

Having established noise model and a prior for transient displacements, we calculate the transient strain rates in the Puget Sound region from eq. (\ref{eq:StrainRate}) over the time period 2010-01-01 to 2017-05-15. Maybe this is the best place to mention the prior for editing the data. Our solution is shown in Figures X through X. Figures X through X show the estimates $\dot{\bm{\epsilon}}$ complete with uncertainties at select times over the course of the Winter 2016 SSE. Figure X, which shows strain rate at the height of the SSE, makes the implications of SSEs for seismic hazard clear.  The SSE causes compression in the Olympic Peninsula and extension in East of Puget Sound. The SSE are thus concentrating strain energy trench-ward, and pushing the subduction zone closer to failure. This is not novel insight because the same conclusions has been drawn from fault slip models which show that SSEs tend to be just at the transition between locked and decoupled. Slow slip event depths from \citet{Dragert2001}, \citet{Wech2009}, \citet{Schmidt2010}, \citet{Bartlow2011} show slip concentrated at depths from 30 to 50 km detph. What is novel is that our estimates of strain rate from SSEs are nonparametric and less likely to be biased by systematic errors such as errors in the fault model. Are uncertainties are also likely to be more conservative since we do not make any dubious assumptions. Thus we can be more confident in our assessment of how much strain is accumulating from SSEs.

>>>>

\section{Method}\label{sec:Method}
In this section, we describe how GNSS displacement observations are used to identify transient crustal strain rates, which we denote as $\dot{\bm{\varepsilon}}(z)$, where $z$ is the ordered pair $(x,t)$, $x$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We consider $\dot{\bm{\varepsilon}}$ to be spatially and temporally coherent deviations from the steady rate of strain accumulation from plate tectonics. We determine $\dot{\bm{\varepsilon}}$ by first identifying transient displacements, $\bm{u}(z)$, which we then spatially and temporally differentiate.  As we will show in Section X, estimates of $\dot{\bm{\varepsilon}}$ turn out to be more effective at illuminating geophysical signal than estimates of $\bm{u}$ or $\dot{\bm{u}}$.  We make a prior assumption that each component of $\bm{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(z) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(z,z')$ is a covariance function indicating how we expect $u_i(z)$ to covary with $u_i(z')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We then drop the component subscripts with the understanding that the same analysis is being repeated to estimate the east, north, and vertical components of $\bm{u}$. We further assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((x,t),(x',t')\right) = \alpha^2 X(x,x')T(t,t'),
\end{equation}  
where $\alpha$ is a hyperparameter. The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in the next section.

@@

Our method for detecting outliers is based on the data editing algorithm described in \citet{Gibbs2011}. We calculate the residuals between the observations and a best fitting model, and data with residuals that are anomalously large are identified as outliers. We again treat $\bm{d}_*$ as a sample of $\bm{d}$. The best fitting model for $\bm{d}_*$ is considered to be the expected value of the random vector $u(\bm{y}) + \bm{P}\bm{a}$ after conditioning it with non-outlier observations. We let $\epsilon$ be uncorrelated noise which has standard deviation given by the formal data uncertainties, $\bm{\sigma}$.  We still consider $u$ to have a separable covariance function as in eq. (\ref{eq:TransientCovariance}), and the choice for $X$ and $T$ does not need to be the same as that used in Section X. Since outliers are determined based on how well a spatially and temporally dependent model fits the data, we are able to identify anomalous observations which may not be immediately apparent based on inspection of individual timeseries.

To begin the algorithm, we let $\Omega$ be the index set of non-outliers in $\bm{d}_*$ and initiate it with all $n$ indices. This algorithm is iterative, and for each iteration we calculate the residual vector
\begin{align}\label{eq:Residual}
\bm{r} &= \frac{\bm{d}_* - \mathrm{E}\left[(u(\bm{y}) + \bm{P}\bm{a})|\tilde{\bm{d}}_* \right]}{\bm{\sigma}} \\
       &= \frac{1}{\bm{\sigma}}\left(\bm{d}_*  - 
\left[ 
\begin{array}{cc}
C_u(\bm{y},\tilde{\bm{y}}) & \bm{P} \\
\end{array}
\right]
\left[
\begin{array}{cc}
C_u(\tilde{\bm{y}},\tilde{\bm{y}}) + \mathrm{diag}(\tilde{\bm{\sigma}}^2) & \tilde{\bm{P}} \\
\tilde{\bm{P}}^T  & \bm{0} \\
\end{array}
\right]^{-1}
\left[
\begin{array}{c}
\tilde{\bm{d}}_* \\
\bm{0} \\
\end{array}
\right] 
\right),
\end{align}
where the tilde indicates that only elements corresponding to indices in $\Omega$ are retained (e.g., $\tilde{\bm{y}} = \{y_i\}_{i\in\Omega}$). We then update $\Omega$ to be
\begin{equation}\label{eq:Update}
\Omega = \{i : |r_i| < \eta \tilde{\bm{r}}_{rms}\}, \ \ \ r_i \in \bm{r}
\end{equation} 
where $\tilde{\bm{r}}_{rms}$ is the root-mean-square of $\tilde{\bm{r}}$ and $\eta$ is an outlier tolerance. We use $\eta=4$ in this study, which seems to accurately identify outliers without unnecessarily decimating the data. Iterations continue until the new $\Omega$ is equal to the previous $\Omega$.

@@

\section{Application Cascadia Slow Slip Events}\label{sec:Cascadia}
In this section we estimate transient strain from GNSS data in Cascadia. We are mainly interested in identifying transient strain resulting from slow slip events (SSEs) \citep[e.g.,][]{Dragert2001}. We show that Gaussian processes are an effective tool for (1) detecting SSEs and (2) understanding how the strain from SSEs are contributing to the strain accumulated over the seismic cycle. Our motivation for detecting SSEs is, in part, because there have been several instances where SSEs immediately preceded megathrust earthquakes \citep{Roeloffs2006}.  In Cascadia, SSEs can be detected by monitoring for associated seismic tremor, which is actively being done by the Pacific Northwest Seismic Network (PNSN) \citep{Wech2010}. 
Here we demonstrate that strain rates derived from GNSS data can also be an effective tool for detecting SSEs. Furthermore, estimated strain rates provide a clear picture of how SSEs transfer elastic strain energy through the crust, which is also important information for assessing seismic hazard. We envision that GNSS derived transient strain rates could be useful for monitoring subduction zones which have SSEs but no associated tremor \citep{Schwartz2007}.

In this study, we use GNSS data which has been made publicly available by the University Navstar Consortium (UNAVCO) \citep{Herring2016} and can be found at www.unavco.org.  We limit the dataset to the stations and time ranges which are pertinent to seven of the most recent SSEs in the Puget Sound region; the earliest SSE began in the August 2010, and the most recent SSE began in the February of 2017. We consider these most recent SSEs because the station coverage is sufficiently dense for us to make well constrained inferences of optimal prior models.  The distribution of stations considered in this station is shown in Figure X.

@@

To determine the appropriate hyperparameters for the FOGM model, $\alpha$ and $\beta$, we first select 38 stations in Washington and Oregon which are east of $121^\circ$ W, where we can reasonably expect no transient signal from SSEs (i.e. $u = 0$). We remove jumps and clean the data with the outlier detection algorithm described in section X.  We then find $\alpha$ and $\beta$ for each station timeseries with the Restricted Maximum Likelihood (REML) method \cite[e.g.,][]{Harville1974,Cressie1992} and Hines 2017. The REML method finds the hyperparameters which maximize the probability of drawing $\bm{d}_*$ from $\bm{d}$. We use the REML method over the maximum likelihood (ML) method \citep[e.g.,][]{Langbein1997} because the REML method accounts for the improper prior that we assigned to $\bm{a}$ without biasing the hyperparameter estimates. The distribution of inferred $\alpha$ and $\beta$ are shown in Figure X. The amplitude of FOGM noise for the east and north components is notable low and are clustered around 0.5 mm/yr$^{0.5}$. The corresponding estimates of $\alpha$ tend to cluster around 0 yr$^{-1}$, suggesting that Brownian motion would have also been an appropriate noise model.  The amplitude of FOGM noise for the vertical component is significantly larger with a median value of 13.8 mm/yr$^{0.5}$. The inferred values for alpha are higher for the vertical component with a median value of 1.87 yr$^{-1}$. In Figure X, we use the median values of $\alpha$ and $\beta$ to generate two random samples of FOGM noise. The samples span seven years and over these seven years the east and north samples drift by about 1 mm. In the context of detecting SSEs, which produce several mm's of surface displacement on the timescale of weeks, the estimated FOGM noise for the east and north component is negligible. In contrast, the estimated FOGM noise for the vertical component is significantly larger than the signal we would expect from SSEs. We suspect that the higher amplitude for the FOGM model in the vertical component is accommodating for deficiencies in our rather simple seasonal model. Based on this analysis, we henceforth ignore temporally correlated noise in the east and north component because of its low amplitude, and we do not estimate $u$ for vertical deformation because of the low signal to noise ratio.

\begin{figure*}
\includegraphics{figures/noise/noise-params.pdf}
\caption{Distribution of estimated FOGM hyperparameters from eq. (\ref{eq:FOGM}).  Hyperparameters are independently estimated for each of the stations shown in Figure X and for each displacement component. ``IQR'' is the inter-quartile range.}   
\label{fig:NoiseParams}
\end{figure*}

\begin{figure*}
\includegraphics{figures/noise/noise-samples.pdf}
\caption{Two FOGM noise samples, where the hyperparameters have been set to the median values from Figure \ref{fig:NoiseParams}.}   
\label{fig:NoiseSamples}
\end{figure*}

@@

\subsection{Transient displacement model}
We next discuss our prior model for transient displacements, and, specifically, our choice for the covariance functions $X(x,x')$ and $T(t,t')$. For $X$, we use the squared exponential (SE) covariance function,
\begin{equation}\label{eq:SE}
X(x,x') = \exp\left(\frac{-||x - x'||_2^2}{2 \beta^2}\right).
\end{equation}
The SE covariance function is commonly used in Kriging \citep[e.g,][]{Cressie1992} and Gaussian process regression \citep[e.g.,][]{Rasmussen2006}.  The SE is a valid (i.e. positive definite) covariance function for any number of spatial dimensions, and it describes an isotropic Gaussian processes with realizations that are infinitely differentiable. In terms of geodetic applications, \citet{Kato1998} and \cite{El-Fiky1999} demonstrated that the SE is an appropriate covariance model for describing long-term tectonic strain rates in Japan.

We consider three potential covariance functions to describe the temporal covariance of $u$.  First, we consider the one-dimensional SE covariance function, simply because the SE describes a good general purpose Gaussian process. Second, we consider integrated Brownian motion (IBM). IBM has zero mean and its covariance function can be found by integrating the covariance function for Brownian motion as
\begin{align}\label{eq:IBM}
T(t,t') &= \int_0^t \int_0^{t'} \min(\tau,\tau') \,d\tau'\,d\tau \\
        &= \frac{1}{2}\min(t,t')^2 \left(\max(t,t') - \frac{1}{3}\min(t,t')\right), \ \ \ t,t' \geq 0.
\end{align}
IBM has been used, in the context of Kalman filtering, as a nonparametric model for the time dependence of geophysical signals \citep[e.g.,][]{Segall1997,McGuire2003,Ohtani2010,Hines2016}. It should be noted that IBM has a reference time at which the Gaussian process is exactly zero. For some geophysical signals, such as postseismic deformation, it is appropriate to have this reference time. However, if we are interesting in detecting transient events, where there is no known start time, then the IBM model may not be appropriate and an isotropic Gaussian process should be preferred. Alternatively we can assume that the start date for the IBM process is long before the first data obvservation; however this can introduce numerical difficulties because the covariance matrix will become ill-conditioned. Third, we consider the Wendland class of covariance functions \citep{Wendland2005}. Wendland functions are positive definite in $\mathbb{R}^d$, and they describes an isotropic Gaussian process with realizations that can be differentiated $k$ times. The form of the covariance function depends on the choice of $d$ and $k$. Wendland covariance functions have compact support, which is a particularly useful property because we can exploit the sparsity of the the corresponding covariance matrices. To describe the temporal covariance of $u$, we only require that $d=1$ and $k=1$. The corresponding Wendland covariance function is 
\begin{equation}\label{eq:Wendland}
T(t,t') = \left(1 - \frac{|t - t'|}{\beta}\right)^3_+ \left(\frac{3|t - t'|}{\beta} + 1\right), 
\end{equation}
where
\begin{equation}
(t)_+ = 
\begin{cases}
t, \ \ \ t > 0 \\
0, \ \ \ \mathrm{otherwise}.
\end{cases}
\end{equation}

We next want to determine appropriate hyperparameters for $X$ and each of the three candidate functions for $T$. We use the REML method to find the optimal hyperparameters for describing each of the seven SSEs. Before using the REML method, we clean the GNSS datasets by removing offsets at times of equipment changes and removing outliers with the method describe in Section X. For the outlier detection algorithm, our prior model, $u$, is chosen to be have a length-scale and time-scale which is sufficiently short to describe SSE displacements. We use the Squared Exponential covariance function for $X$ with $\beta = 100$ km, and we use the Wendland covariance function for $T$, due to its computational efficiency, with $\beta = 0.1$ yr$^{-1}$. The product of the two covariance functions is then scaled by $\alpha^2 = 1$ mm$^{2}$. The outlier detection algorithm is particularly effective at removing outliers for stations at high elevation, which can be adversely affected by ice or snow \citep{Lisowski2008}. As an example, Figure \ref{fig:Outliers} shows the outliers which were automatically detected for station SC03, which is located on Mount Olympus in Washington. Note that most of the detected outliers occur during the winter and can likely be attributed to ice or snow.  After cleaning the dataset, we divide it into four month long subsets, where each subset is roughly centered on the time of an SSE.  We identify the times of seven SSEs using tremor records from \cite{Wech2010}. For each data subset, we use the REML method to find the hyperparameters for $T$ and $X$ that best describe deformation from the SSE. Our motivation analyzing subsets of the data is two-fold: (1) it keeps the analysis computationally tractable and (2) the distribution of estimated hyperparameters will indicates whether we can use a single covariance model to describe all SSEs. Our hyperparameter estimates are summarized in Table X.

\begin{figure*}
\includegraphics{figures/outliers/outliers.pdf}
\caption{Detrended easting component of GNSS displacement data for station SC03, which is located on Mt. Olympus in Washington.  The orange markers indicate outliers which were automatically detected using the algorithm from Section \ref{sec:Method}. The error bars show one standard deviation uncertainty.}   
\label{fig:Outliers}
\end{figure*}

We next want to identify which covariance function for $T$ best describes the SSEs. We could use the REML likelihoods to compare different models, similar to \citet{Langbein2004}. However, the maximum likelihoods do not vary by more than a factor of X between the models, which does not definitively identify the best model. We then compare the posterior estimates $\hat{u}$ for each model to the observations to see if any model is noticeably better at describing the data. We show $\hat{u}$ using the three different function for $T$ in Figure X. It is clear that neither model offer any distinct improvement over the others. For most stations $\hat{u}$ appears to accurately capture the deformation from SSEs; however, for some stations which are closer to Puget Sound such as X and X, the rate of deformation is underestimated by the best fitting model. We could have manually specified a prior. We choose the Wendland function to describe $T$ because evaluating $T$ will yield sparse covariance matrices. We utilize the sparsity of the covariance matrices with the sparse matrix library CHOLMOD \citep{Chen2008}, which significantly expands the amount of data which we are able to tractably analyze with the method described in Section X.

\subsection{Transient Strain Rates} 
Having established noise model and a prior for transient displacements, we use cleaned GNSS dataset described in the previous section to calculate transient strain rates in the Puget Sound region from eq. (\ref{eq:StrainRate}). Our solution is shown in Figures X through X. Figures X through X show the estimates $\dot{\bm{\epsilon}}$ complete with uncertainties at select times over the course of the Winter 2016 SSE. Figure X, which shows strain rate at the height of the SSE, makes the implications of SSEs for seismic hazard clear.  The SSE causes compression in the Olympic Peninsula and extension in East of Puget Sound. The SSE are thus concentrating strain energy trench-ward, and pushing the subduction zone closer to failure. This is not novel insight because the same conclusions has been drawn from fault slip models which show that SSEs tend to be just at the transition between locked and decoupled. Slow slip event depths from \citet{Dragert2001}, \citet{Wech2009}, \citet{Schmidt2010}, \citet{Bartlow2011} show slip concentrated at depths from 30 to 50 km detph. What is novel is that our estimates of strain rate from SSEs are nonparametric and less likely to be biased by systematic errors such as errors in the fault model. Are uncertainties are also likely to be more conservative since we do not make any dubious assumptions. Thus we can be more confident in our assessment of how much strain is accumulating from SSEs.
